Cochain level May--Steenrod operations

### 2021-08-20 aux/abstract.tex
Structural refactoring using new template

\begin{abstract}
	Steenrod defined in 1947 the Steenrod squares on the mod 2 cohomology of spaces using explicit cochain formulae for the cup-$i$ products; a family of coherent homotopies derived from the broken symmetry of Alexander--Whitney's chain approximation to the diagonal. He later defined his homonymous operations for all primes using the homology of symmetric groups. This approach enhanced the conceptual understanding of the operations and allowed for many advances, but lacked the concreteness of their definition at the even prime. In recent years, thanks to the development of new applications of cohomology, having definitions of Steenrod operations that can be effectively computed in specific examples has become a key issue. This article provides such definitions at all primes using the operadic viewpoint of May, and defines multioperations that generalize the cup-$i$ products of Steenrod on the simplicial and cubical cochains of spaces.
\end{abstract}

>>>>

\begin{abstract}
	Steenrod defined in 1947 the Steenrod squares on the mod 2 cohomology of spaces using explicit cochain formulae for the cup-$i$ products; a family of coherent homotopies derived from the broken symmetry of Alexander--Whitney's chain approximation to the diagonal. He later defined his homonymous operations for all primes using the homology of symmetric groups. This approach enhanced the conceptual understanding of the operations and allowed for many advances, but lacked the concreteness of their definition at the even prime. In recent years, thanks to the development of new applications of cohomology,the need to have an effectively computable definition of Steenrod operations has become a key issue. This article provides such definitions at all primes using the operadic viewpoint of May, and defines multioperations that generalize the cup-$i$ products of Steenrod on the simplicial and cubical cochains of spaces.
\end{abstract}

### 2021-08-20 may_st.tex
Structural refactoring using new template

\documentclass{amsart}
\usepackage{amssymb}
\usepackage{tikz-cd}
\usepackage{graphics}
% hyperref
\usepackage[bookmarks=true, linktocpage=true,
bookmarksnumbered=true, breaklinks=true,
pdfstartview=FitH, hyperfigures=false,
plainpages=false, naturalnames=true,
colorlinks=true, pagebackref=true,
pdfpagelabels]{hyperref}
\hypersetup{
	colorlinks,
	citecolor=blue,
	filecolor=blue,
	linkcolor=blue,
	urlcolor=blue
}
\usepackage[capitalize, noabbrev]{cleveref}
\usepackage[alphabetic]{amsrefs}

% enviorenments
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

% sets
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\F}{\mathbb{F}}

% funtors and operators
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\End}{\mathrm{End}}
\DeclareMathOperator*{\colim}{colim}

% other
\renewcommand{\th}{^{\mathrm{th}}}

% paper shape
\setlength{\textwidth}{\paperwidth}
\addtolength{\textwidth}{-1.6in}
\setlength{\textheight}{\paperheight}
\addtolength{\textheight}{-1.8in}
\calclayout

\include{figures}

\begin{document}

@@

\input{abstract}

\maketitle

\input{introduction}
\input{preliminaries}
\input{groups}
\input{operads}
\input{steenrod}
\input{effective}
\input{outlook}

\bibliography{bibliography}

\end{document}

>>>>

\documentclass{amsart}
\include{aux/style}
\include{aux/usualcmds}

@@

\include{aux/figures}

\begin{document}
	\input{aux/abstract}
	\maketitle
	\input{sec/introduction}
	\input{sec/preliminaries}
	\input{sec/groups}
	\input{sec/operads}
	\input{sec/steenrod}
	\input{sec/effective}
	\input{sec/outlook}
	\printbibliography
\end{document}

### 2021-08-20 sec/effective.tex
Structural refactoring using new template

In this section we construct explicit May--Steenrod structures on three well know combinatorial $E_\infty$-operads: the Barratt--Eccles operad $\mathcal E$ \cite{berger04combinatorial}, the surjection operad $\mathcal X$ \cite{mcclure03cochain}, and the operad $U(\mathcal M)$ associated to the finitely presented $E_\infty$-prop $\mathcal M$ introduced in \cite{medina2020prop1}.
We also define natural and effective May--Steenrod structure on the normalized cochains of any simplicial or cubical set using that these are algebras over the operad $U(\mathcal M)$.

@@

We begin by reviewing the $\mathrm{S}$-module structure underlying the Barratt--Eccles operad and, since will not use it in this work, refer to \cite{berger04combinatorial} for a description of if composition structure.
For a non-negative integer $r$ define the simplicial set $E(\mathrm S_r)$ by
\begin{equation} \label{eq: milnor model of symmetric}
\begin{split}
E(\mathrm S_r)_n &= \{ (\sigma_0, \dots, \sigma_n)\ |\ \sigma_i \in \mathrm{S}_r\}, \\
d_i(\sigma_0, \dots, \sigma_n) &= (\sigma_0, \dots, \widehat{\sigma}_i, \dots, \sigma_n), \\
s_i(\sigma_0, \dots, \sigma_n) &= (\sigma_0, \dots, \sigma_i, \sigma_i, \dots, \sigma_n) \\
\end{split}
\end{equation}
with a left $\mathrm S_r$-action given by
\begin{equation*}
\sigma (\sigma_0, \dots, \sigma_n) = (\sigma \sigma_0, \dots, \sigma \sigma_n).
\end{equation*}
The chain complex resulting from applying the functor of normalized integral chains to it is denoted $\mathcal E(r)$, which corresponds to the arity $r$ part of the Barratt--Eccles operad.

@@

In this subsection we introduce a May--Steenrod structures on the surjection operad.
There are two widely used sign conventions for this operad respectively due to McClure--Smith \cite{mcclure03cochain} and Berger--Fresse \cite{berger04combinatorial}.
Using the Berger--Fresse convention, we can define a May--Steenrod structure on the surjection operad by composing the map $\psi_{\mathcal E}$ with the \textit{table reduction} quasi-isomorphism $\mathcal E \to \mathcal X$ introduced in Section 1.3 of \cite{berger04combinatorial}.
We define an May--Steenrod structure on the surjection operad in a convention independent way which recovers the table reduction May--Steenrod structure in the Berger--Fresse case.

Let us start by recalling the definition of the surjection operad.
For a non-negative integer $r$ let $\mathcal X(r)_n$ be the free abelian group generated by all functions from $\{1, \dots, n+r\}$ to $\{1, \dots, r\}$ modulo the subgroup generated by degenerate functions, i.e., those which are either non-surjective or have a pair of equal consecutive values.
We only describe the McClure--Smith convention since we refer to it in subsequent sections.
The boundary map and symmetric action in this case is defined using the Koszul convention regarding a surjection $s \colon \{1, \dots, n+r\} \to \{1, \dots, r\} $ as the top dimensional generator in the chain complex
\begin{equation} \label{eq: surjection boundary}
\bigotimes_{i = 1}^r N_\bullet \big( \triangle^{s^{-1}(i)}; \Z \big).
\end{equation}
Explicitly, if we think of $s$ as a sequence of integers $\big( s(1), \dots, s(n+r) \big)$ the boundary of $s$ is the sum of sequences obtained by orderly removing one at the time the occurrences of $1$ with alternating signs, then those of $2$ with starting sign equal to that of the last removal of $1$, and so on.
Since we do not use the composition structure we refer to \cite{mcclure03cochain} for it.

@@

We will define a May--Steenrod structure $\psi_{U(\mathcal M)} \colon \mathcal W \to U(\mathcal M)$ by composing the May--Steenrod structure $\psi_{\mathcal X} \colon \mathcal W \to \mathcal X$ with a morphism $SL \colon \mathcal X \to U(\mathcal M)$ of $\mathrm{C}$-modules ($\mathrm{S}$-modules in fact) that we now define.
Given a surjection $s \colon \{1, \dots, n+r\} \to \{1, \dots, r\}$ let $SL(s)$ be the element represented by the immersed connected $(1,r)$-graph
\begin{center}
	\begin{tikzpicture}[scale=.85]
	\draw (0,0)--(0,-.6) node[below]{$\scriptstyle 1$};
	\draw (0,0)--(.5,.5);
	\draw (-.3, .3)-- (-.2,.5) node[above]{\quad $\scriptstyle 1\ 2\, ...\, k_1$};
	\draw (-.5,.5)--(0,0);
	\node at (.11,.4){$...$};

	\node at (1,0){$\scriptstyle \cdots$};
	\node at (1,-.9){$\scriptstyle \cdots$};

	\draw (2,0)--(2,-.68) node[below]{$\scriptstyle r$};
	\draw (2,0)--(2.5,.5);
	\draw (1.7, .3)-- (1.8,.5) node[above]{\quad $\scriptstyle 1\ 2\, ...\, k_r$};
	\draw (1.5,.5)--(2,0);
	\node at (2.11,.4){$...$};

	\draw (1,2.5)--(1,3) node[above]{$\scriptstyle 1$};
	\draw (1,2.5)--(0,2) node[below]{$\scriptstyle 1$};
	\draw (.25,2.125)--(.5,2) node[below]{$\scriptstyle 2$};
	\draw (.5,2.25)--(1,2) node[below]{$\scriptstyle 3$};
	\draw (1,2.5)--(2,2) node[below]{\quad $\scriptstyle n + r$};
	\node at (1.5,1.75){$...$};

	\node at (1,1.3) {$\vdots$};

	\node at (2.85,0){};
	\end{tikzpicture}
\end{center}
that has no internal vertices and such that the $n+r$ strands at the top are orderly connected to the strands at the bottom following the values of $s$.

@@

\begin{example}
	Figure \ref{fig: cup 0,1,2} shows $\psi_{U(\mathcal M)}(2)(e_n)$ for small values of $n$.
\end{example}

\begin{figure}[h]
	\centering
	\boxed{
		\begin{tikzpicture}[scale=.55]
		\draw (1,3.7) to (1,3);

		\draw (1,3) to [out=205, in=90] (0,0);
		\draw (1,3) to [out=-25, in=90] (2,0);

		\node at (1,-.5){$n=0$};
		\end{tikzpicture}\hspace*{1cm}
		\begin{tikzpicture}[scale=.55]
		\draw (1,3.7) to (1,3);

		\draw (1,3) to [out=205, in=90] (0,0);

		\draw [shorten >= 0cm] (.6,2.73) to [out=-100, in=90] (2,0);
		%\draw [shorten <= 0cm] (1,1.5) to [out=-60, in=90] (2,0);

		\draw [shorten >= .15cm] (1,3) to [out=-25, in=30, distance=1.1cm] (1,1.5);
		\draw [shorten <= .1cm] (1,1.5) to [out=210, in=20] (0,1);

		\node at (1,-.5){$n=1$};
		\end{tikzpicture}\hspace*{1cm}
		\begin{tikzpicture}[scale=.55]
		\draw (1,3.7) to (1,3);

		\draw (1,3) to [out=205, in=90] (0,0);
		\draw (1,3) to [out=-25, in=90] (2,0);

		\draw [shorten >= 0cm] (.6,2.73) to [out=210, in=135] (1,1.5);
		\draw [shorten <= 0cm] (1,1.5) to [out=-45, in=170] (2,1);

		\draw [shorten >= .1cm] (1.4,2.73) to [out=-30, in=45] (1,1.5);
		\draw [shorten <= .1cm] (1,1.5) to [out=-135, in=10] (0,1);

		\node at (1,-.5){$n=2$};
		\end{tikzpicture}
	}
	\caption{The values of $\psi_{U(\mathcal M)}(2)(e_n)$ for small values of $n$.}
	\label{fig: cup 0,1,2}
\end{figure}

@@

\begin{remark}
	The $E_\infty$-structure we described in Proposition~\ref{prop: simplicial chain bialgebra}, depending solely on three fundamental maps, generalizes the coalgebra structures of McClure--Smith \cite{mcclure03cochain} and Berger--Fresse \cite{berger04combinatorial}, please consult \cite{medina2020prop1} for more details.
\end{remark}

@@

\begin{example}
	Let us consider the prime $3$.
	The value of $\beta P^0(x)\big([01]^2\big)$ for a homogeneous cocycle $x$ in $N^{-1}(\square^2)$ is equal to the value of $x^{\otimes 3}$ acting on
	\begin{align*}
	\,-\, [01]1 \otimes [01]0 \otimes 1[01] \,-\, 0[01] \otimes [01]0 \otimes 1[01] + [01]1 \otimes 0[01] \otimes [01]1 + 0[01] \otimes 0[01] \otimes [01]1.
	\end{align*}
\end{example}

>>>>

In this section we construct explicit May--Steenrod structures on three well know combinatorial $E_\infty$-operads: the Barratt--Eccles operad $\mathcal E$ \cite{berger2004combinatorial}, the surjection operad $\mathcal X$ \cite{mcclure2003multivariable}, and the operad $U(\mathcal M)$ associated to the finitely presented $E_\infty$-prop $\mathcal M$ introduced in \cite{medina2020prop1}.
We also define natural and effective May--Steenrod structure on the normalized cochains of any simplicial or cubical set using that these are algebras over the operad $U(\mathcal M)$.

@@

We begin by reviewing the $\mathrm{S}$-module structure underlying the Barratt--Eccles operad and, since will not use it in this work, refer to \cite{berger2004combinatorial} for a description of if composition structure.
For a non-negative integer $r$ define the simplicial set $E(\mathrm S_r)$ by
\begin{equation} \label{eq: milnor model of symmetric}
\begin{split}
E(\mathrm S_r)_n &= \{ (\sigma_0, \dots, \sigma_n)\ |\ \sigma_i \in \mathrm{S}_r\}, \\
d_i(\sigma_0, \dots, \sigma_n) &= (\sigma_0, \dots, \widehat{\sigma}_i, \dots, \sigma_n), \\
s_i(\sigma_0, \dots, \sigma_n) &= (\sigma_0, \dots, \sigma_i, \sigma_i, \dots, \sigma_n) \\
\end{split}
\end{equation}
with a left $\mathrm S_r$-action given by
\begin{equation*}
\sigma (\sigma_0, \dots, \sigma_n) = (\sigma \sigma_0, \dots, \sigma \sigma_n).
\end{equation*}
The chain complex resulting from applying the functor of normalized integral chains to it is denoted $\mathcal E(r)$, which corresponds to the arity $r$ part of the Barratt--Eccles operad.

@@

In this subsection we introduce a May--Steenrod structures on the surjection operad.
There are two widely used sign conventions for this operad respectively due to McClure--Smith \cite{mcclure2003multivariable} and Berger--Fresse \cite{berger2004combinatorial}.
Using the Berger--Fresse convention, we can define a May--Steenrod structure on the surjection operad by composing the map $\psi_{\mathcal E}$ with the \textit{table reduction} quasi-isomorphism $\mathcal E \to \mathcal X$ introduced in Section 1.3 of \cite{berger2004combinatorial}.
We define an May--Steenrod structure on the surjection operad in a convention independent way which recovers the table reduction May--Steenrod structure in the Berger--Fresse case.

Let us start by recalling the definition of the surjection operad.
For a non-negative integer $r$ let $\mathcal X(r)_n$ be the free abelian group generated by all functions from $\{1, \dots, n+r\}$ to $\{1, \dots, r\}$ modulo the subgroup generated by degenerate functions, i.e., those which are either non-surjective or have a pair of equal consecutive values.
We only describe the McClure--Smith convention since we refer to it in subsequent sections.
The boundary map and symmetric action in this case is defined using the Koszul convention regarding a surjection $s \colon \{1, \dots, n+r\} \to \{1, \dots, r\} $ as the top dimensional generator in the chain complex
\begin{equation} \label{eq: surjection boundary}
\bigotimes_{i = 1}^r N_\bullet \big( \triangle^{s^{-1}(i)}; \Z \big).
\end{equation}
Explicitly, if we think of $s$ as a sequence of integers $\big( s(1), \dots, s(n+r) \big)$ the boundary of $s$ is the sum of sequences obtained by orderly removing one at the time the occurrences of $1$ with alternating signs, then those of $2$ with starting sign equal to that of the last removal of $1$, and so on.
Since we do not use the composition structure we refer to \cite{mcclure2003multivariable} for it.

@@

We will define a May--Steenrod structure $\psi_{U(\mathcal M)} \colon \mathcal W \to U(\mathcal M)$ by composing the May--Steenrod structure $\psi_{\mathcal X} \colon \mathcal W \to \mathcal X$ with a morphism $SL \colon \mathcal X \to U(\mathcal M)$ of $\mathrm{C}$-modules ($\mathrm{S}$-modules in fact) that we now define.
Given a surjection $s \colon \{1, \dots, n+r\} \to \{1, \dots, r\}$ let $SL(s)$ be the element represented by the immersed connected $(1,r)$-graph

\begin{center}
	\input{aux/surjection_like}
\end{center}

that has no internal vertices and such that the $n+r$ strands at the top are orderly connected to the strands at the bottom following the values of $s$.

@@

\begin{example}
	The following immersed $(1,2)$-graphs are the elements $\psi_{U(\mathcal M)}(2)(e_n)$ for small values of $n$:
	\begin{center}
		\input{aux/cupi_graphs}
	\end{center}
\end{example}

@@

\begin{remark}
	The $E_\infty$-structure we described in Proposition~\ref{prop: simplicial chain bialgebra}, depending solely on three fundamental maps, generalizes the coalgebra structures of McClure--Smith \cite{mcclure2003multivariable} and Berger--Fresse \cite{berger2004combinatorial}, please consult \cite{medina2020prop1} for more details.
\end{remark}

@@

\begin{example}
	Let us consider the prime $3$.
	The value of $\beta P^0(x)\big([01]^2\big)$ for a homogeneous cocycle $x$ in $N^{-1}(\square^2)$ is equal to the value of $x^{\otimes 3}$ acting on
	\begin{align*}
	&\,-\, [01]1 \otimes [01]0 \otimes 1[01] \,-\, 0[01] \otimes [01]0 \otimes 1[01] \\
	&\,+\, [01]1 \otimes 0[01] \otimes [01]1 + 0[01] \otimes 0[01] \otimes [01]1.
	\end{align*}
\end{example}

### 2021-08-20 sec/groups.tex
Structural refactoring using new template

The homology of $\mathrm{S}_r$ is harder to compute.
With untwisted coefficients, the method of computation followed by several authors was to prove the injectivity of this homology into that of the infinite symmetric group and take advantage of a natural Hopf algebra structure on it.
A powerful result stemming from the deep connection of this question with infinite loop space theory, is the existence of a homology isomorphism of spaces
\begin{equation*}
\Z \times B\mathrm{S}_\infty \to Q(S^0) = \Omega^\infty \Sigma^\infty (S^0)
\end{equation*}
credited to Dyer--Lashof \cite{dyer62lashof}, Barratt--Priddy and Quillen \cite{barratt1972priddy}.

@@

\begin{proof}
	This is proven as Theorem 4.1 in \cite{steenrod53cyclic} where Thom is also credited with a different proof.
\end{proof}

>>>>

The homology of $\mathrm{S}_r$ is harder to compute.
With untwisted coefficients, the method of computation followed by several authors was to prove the injectivity of this homology into that of the infinite symmetric group and take advantage of a natural Hopf algebra structure on it.
A powerful result stemming from the deep connection of this question with infinite loop space theory, is the existence of a homology isomorphism of spaces
\begin{equation*}
\Z \times B\mathrm{S}_\infty \to Q(S^0) = \Omega^\infty \Sigma^\infty (S^0)
\end{equation*}
credited to Dyer--Lashof \cite{dyer62lashof}, Barratt--Priddy and Quillen \cite{barratt1972priddyquillen}.

@@

\begin{proof}
	This is proven as Theorem 4.1 in \cite{steenrod1953cyclic} where Thom is also credited with a different proof.
\end{proof}

### 2021-08-20 sec/introduction.tex
Structural refactoring using new template

The role Steenrod operations play in Stable Homotopy Theory is hard to overstate.
The reason is that, given the representability of the cohomology functor, these operations together with the Bockstein homomorphism can be used to give a complete description of the algebraic structure naturally present on the mop-$p$ \mbox{cohomology} algebra of spaces.
For the even prime, Steenrod squares were introduced in \cite{steenrod47products} via an explicit choice of coherent homotopical corrections to the broken symmetry of Alexander--Whitney's chain approximation to the diagonal, the so called cup-$i$ products.
Later, for odd primes, their definition was given non-effectively using arguments based on the mod $p$ homology of symmetric groups \cites{steenrod53symmetric, steenrod53cyclic, steenrod62operations}.
This viewpoint enhanced the conceptual understanding of the operations and allowed for many advances \cites{adem1952iteration, milnor1958dual, adams1995stable}, but lacked the concreteness of their definition at the even prime.
The purpose of this paper is to fill this gap in the literature, introducing effective descriptions of multioperations at the cochain level generalizing Steenrod's cup-$i$ products defining Steenrod operations at all primes.

In recent years, thanks to the development of new applications of cohomology --~most notably in Applied Topology and Quantum Field Theory~-- the need to have a definition of Steenrod operations that can be effectively computed in specific examples has gain considerable importance.
In Applied Topology, the use of persistence homology \cites{carlsson2009data, edelsbrunner2008persistent} has open many new interdisciplinary research directions \cites{de2007coverage, chan2013topology, lee2017quantifying}, and the availability of formulae for the cup-$i$ products allowed for the development of a theory of persistence Steenrod modules accessing finer features of the data \cite{medina2018persistence}.
The chain level viewpoint for the study of homology operations was brought to the forefront by String Topology \cite{Sullivanoverview}, and as served as a major driver of innovation \cites{TZ, hoch2}.
In Lattice Field Theory, cup-$i$ products are used in the construction of effective actions \cite{gaiotto2016spin, bhardwaj2017state, kapustin2017fermionic}, and this connection with physics motivated the development of effective versions of spin bordism \cites{brumfiel2016pontrjagin, brumfiel2018pontrjagin} prominently featuring higher derived structures at the cochain level \cite{medina2020cartan, medina2020adem}.

Following \cite{may70generalapproach}, we take a more general approach to Steenrod operations that also includes Araki--Kudo--Dyer--Lashof operations on the mod $p$ homology of infinite loop spaces \cites{araki56squaring, dyer62lashof}.
We use the language of operads \cite{may72geometry} to describe at the (co)chain level the integral structure required to define (co)homology operations at every prime.
We then describe effective constructions of this structure on three prominent models of the $E_\infty$-operad, identifying elements in them that represent Steenrod operations in the mod $p$ homology of their algebras; these are the Barratt--Eccles \cite{berger04combinatorial}, surjection \cite{mcclure03cochain}, and $U(\mathcal M)$ \cite{medina2020prop1} operads.
Since the cochains of simplicial sets are equipped with effective and compatible algebra structures over each of these operads, we are able to explicitly describe canonical multioperations generalizing the \mbox{cup-$i$} products of Steenrod to every prime.
An alternative approach based on the Eilenberg--Zilber contraction can be found in \cite{gonzalez2005hpt}.
The $p = 2$ part of our constructions specialize in the simplicial context to the definition of cup-$i$ products given by Steenrod \cite{steenrod47products} and recovered by several authors \cites{mcclure03cochain, berger04combinatorial, medina2021newformulas}, and in the cubical setting to the cup-$i$ products of Kadeishvili \cite{kadeishvili1998dg} and Kr\v{c}\'{a}l--Pilarczy \cite{pilarczyk2016cubical}.

A context where the cubical viewpoint arises naturally is the study of loop spaces.
This is through Baues' geometric generalization of Adams' cobar construction \cite{baues1980geometry, adams1957cobar}.
Using Baues' work, an application of the constructions presented in this paper is the explicit descriptions of Steenrod operations on the cobar construction on the coalgebra of chains of reduced simplicial sets \cite{medina2021cobar}.

@@

A.M-M. acknowledges financial support from Innosuisse grant \mbox{32875.1 IP-ICT - 1}, and the hospitality of the \textit{Laboratory for Topology and Neuroscience} at EPFL

>>>>

The role Steenrod operations play in stable homotopy theory is hard to overstate.
The reason is that, given the representability of the cohomology functor, these operations together with the Bockstein homomorphism can be used to give a complete description of the algebraic structure naturally present on the mop-$p$ \mbox{cohomology} algebra of spaces.
For the even prime, Steenrod squares were introduced in \cite{steenrod1947products} via an explicit choice of coherent homotopical corrections to the broken symmetry of Alexander--Whitney's chain approximation to the diagonal, the so called cup-$i$ products.
Later, for odd primes, their definition was given non-effectively using arguments based on the mod $p$ homology of symmetric groups \cite{steenrod1952reduced, steenrod1962cohomology, steenrod1962cohomology}.
This viewpoint enhanced the conceptual understanding of the operations and allowed for many advances \cite{adem1952iteration, milnor1958dual, adams1995stable}, but lacked the concreteness of their definition at the even prime.
The purpose of this paper is to fill this gap in the literature, introducing effective descriptions of multioperations at the cochain level generalizing Steenrod's cup-$i$ products defining Steenrod operations at all primes.

In recent years, thanks to the development of new applications of cohomology, most notably in applied topology and topological quantum field theories, the need to have an effectively computable definition of Steenrod operations has gain considerable importance.

In applied topology, the use of persistence homology \cite{edelsbrunner2002topological, carlsson2005barcode} has open many new interdisciplinary research directions \cite{chan2013viral, hess2017cliques}, and the availability of formulae for the cup-$i$ products allowed for the development of a theory of persistence Steenrod modules accessing finer features of the data \cite{medina2018persistence}.

The chain level viewpoint for the study of homology operations was brought to the forefront by string topology \cite{sullivan2007presentstate}, and as served as a major driver of innovation \cite{tradler2007string, hoch2}.

In lattice field theory, cup-$i$ products are used in the construction of effective actions \cite{gaiotto2016spin, bhardwaj2017fermionic, kapustin2017fermionic}, and this connection with physics motivated the development of effective versions of spin bordism \cite{brumfiel2016pontrjagin, brumfiel2018quadratic} prominently featuring higher derived structures at the cochain level \cite{medina2020cartan, medina2021adem}.

Following \cite{may1970general}, we take a more general approach to Steenrod operations that also includes Araki--Kudo--Dyer--Lashof operations on the mod $p$ homology of infinite loop spaces \cite{araki56squaring, dyer62lashof}.
We use the language of operads \cite{may1972geometry} to describe at the (co)chain level the integral structure required to define (co)homology operations at every prime.
We then describe effective constructions of this structure on three prominent models of the $E_\infty$-operad, identifying elements in them that represent Steenrod operations in the mod $p$ homology of their algebras; these are the Barratt--Eccles \cite{berger2004combinatorial}, surjection \cite{mcclure2003multivariable}, and $U(\mathcal M)$ \cite{medina2020prop1} operads.
Since the cochains of simplicial sets are equipped with effective and compatible algebra structures over each of these operads, we are able to explicitly describe canonical multioperations generalizing the \mbox{cup-$i$} products of Steenrod to every prime.
An alternative approach based on the Eilenberg--Zilber contraction can be found in \cite{gonzalez2005cocyclic}.
The $p = 2$ part of our constructions specialize in the simplicial context to the definition of cup-$i$ products given by Steenrod \cite{steenrod1947products} and recovered by several authors \cite{mcclure2003multivariable, berger2004combinatorial, medina2021newformulas}, and in the cubical setting to the cup-$i$ products of Kadeishvili \cite{kadeishvili2003cupi} and Kr\v{c}\'{a}l--Pilarczy \cite{pilarczyk2016cubical}.

A context where the cubical viewpoint arises naturally is the study of loop spaces.
This is through Baues' geometric generalization of Adams' cobar construction \cite{adams1956cobar, baues1998hopf}.
Using Baues' work, an application of the constructions presented in this paper is the explicit descriptions of Steenrod operations on the cobar construction on the coalgebra of chains of reduced simplicial sets \cite{medina2021cobar}.

@@

A.M-M. acknowledges financial support from Innosuisse grant \mbox{32875.1 IP-ICT-1}, and the hospitality of the \textit{Laboratory for Topology and Neuroscience} at EPFL

### 2021-08-20 sec/operads.tex
Structural refactoring using new template

\section{$\Gamma$-modules, operads and props} \label{s:operads}

@@

Succinctly, an operad $\mathcal O$ is an $\mathrm{S}$-module together with a collection of $R$-linear maps
\begin{equation*}
\mathcal O(r) \otimes \mathcal O(s) \to \mathcal O(r+s-1)
\end{equation*}
satisfying suitable associativity, equivariance and unitality conditions.
A prop $\mathcal P$ is an $\mathrm{S}$-bimodule together with two types of compositions; horizontal
\begin{equation*}
\mathcal P(r_1, s_1) \otimes \mathcal P(r_2, s_2) \to \mathcal P(r_1 + r_2, s_1 + s_2)
\end{equation*}
and vertical
\begin{equation*}
\mathcal P(r,s) \otimes \mathcal P(s, t) \to \mathcal P(r, t)
\end{equation*}
satisfying their own versions of associativity, equivariance and unitality.
For a complete presentation of these concepts we refer to Definition 11 and 54 of \cite{markl2008operads}.

@@

\begin{definition} [\cite{may72geometry}, \cite{boardman1973homotopy}] \label{def: e-infinity operad and prop}
	An operad is said to be an $E_\infty$-operad if its underlying \mbox{$\mathrm{S}$-module} is $E_\infty$, and a prop $\mathcal P$ is said to be an $E_\infty$-prop if either $U_1(\mathcal P)$ or $U_2(\mathcal P)$ is an \mbox{$E_\infty$-operad}.
\end{definition}

>>>>

\section{\texorpdfstring{$\Gamma$}{Gamma}-modules, operads and props} \label{s:operads}

@@

Succinctly, an operad $\mathcal O$ is an $\mathrm{S}$-module together with a collection of $R$-linear maps
\begin{equation*}
\mathcal O(r) \otimes \mathcal O(s) \to \mathcal O(r+s-1)
\end{equation*}
satisfying suitable associativity, equivariance and unitality conditions.
A prop $\mathcal P$ is an $\mathrm{S}$-bimodule together with two types of compositions; horizontal
\begin{equation*}
\mathcal P(r_1, s_1) \otimes \mathcal P(r_2, s_2) \to \mathcal P(r_1 + r_2, s_1 + s_2)
\end{equation*}
and vertical
\begin{equation*}
\mathcal P(r,s) \otimes \mathcal P(s, t) \to \mathcal P(r, t)
\end{equation*}
satisfying their own versions of associativity, equivariance and unitality.
For a complete presentation of these concepts we refer to Definition 11 and 54 of \cite{markl2008props}.

@@

\begin{definition} [\cite{may1972geometry}, \cite{boardman1973homotopy}] \label{def: e-infinity operad and prop}
	An operad is said to be an $E_\infty$-operad if its underlying \mbox{$\mathrm{S}$-module} is $E_\infty$, and a prop $\mathcal P$ is said to be an $E_\infty$-prop if either $U_1(\mathcal P)$ or $U_2(\mathcal P)$ is an \mbox{$E_\infty$-operad}.
\end{definition}

### 2021-08-20 sec/outlook.tex
Structural refactoring using new template

In Higher Category Theory, the paper \cite{medina2020globular} constructs a functor producing strict \mbox{$\infty$-cat}egories from group-like cup-$i$ coalgebras in a manner similar to \cite{steiner2004omega}.
In particular, the cup-$i$ constructions described in this article for standard simplices and cubes define, respectively, the Street nerve and cubical nerve of strict $\infty$-categories.
We anticipate that the more general Steenrod $(p,i)$-products constructed in this work will also have deep combinatorial interpretations.

@@

In this article we have not focused on the operations that exist non-trivially for \mbox{$E_n$-algebras} with $n$ finite, see part III of \cite{may76homology}.
A treatment close to ours for the $E_2$ case was given in \cite{Tourtchine}.
Effective constructions for these Dyer--Lashof--Cohen operations should be related to the geometry and combinatorial structure of configuration spaces \cite{KZhang, sinha2013littledisks, berger04combinatorial, ayala2014configuration} and higher categories \cite{Bathigher, BalFiedSchwVogt, Rezkhigher}, and will appear elsewhere.

>>>>

In Higher Category Theory, the paper \cite{medina2020globular} constructs a functor producing strict $\infty$-categories from group-like cup-$i$ coalgebras in a manner similar to \cite{steiner2004omega}.
In particular, the cup-$i$ constructions described in this article for standard simplices and cubes define, respectively, the Street nerve and cubical nerve of strict $\infty$-categories.
We anticipate that the more general Steenrod $(p,i)$-products constructed in this work will also have deep combinatorial interpretations.

@@

In this article we have not focused on the operations that exist non-trivially for \mbox{$E_n$-algebras} with $n$ finite, see part III of \cite{may76homology}.
A treatment close to ours for the $E_2$ case was given in \cite{tourtchine2006cohen}.
Effective constructions for these Dyer--Lashof--Cohen operations should be related to the geometry and combinatorial structure of configuration spaces \cite{KZhang, sinha2013littledisks, berger2004combinatorial, ayala2014configuration} and higher categories \cite{Bathigher, BalFiedSchwVogt, Rezkhigher}, and will appear elsewhere.
